Identifying Shifts in the Share Distribution of Transboundary Fish Stocks to Lead Time for Policy Response Under Climate Change

### 2020-11-16 Scripts/Manuscript.Rmd
Updates to chapter

---
title: Identifying Shifts in the Share Distribution of Transboundary Fish Stocks to Lead Time for Policy Response Under Climate Change
subtitle: Challenges to transboundary fisheries under climate change
author: 'Authors: Juliano Palacios-Abrantes*^1^, Thomas Frölicher^2^, Gabriel Reygondeau^13^, U. Rashid Sumaila^1^, Colette Wabnitz^14^, and William W.L. Cheung^1^'
date: 'Affiliations: ^1^Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada. ^2^Climate and Environmental Physics, Physics Institute, University of Bern.'
output:
  # word_document: 
    # reference_docx: ../References/template.docx
  pdf_document: default
bibliography: ../References/ToE_Ref.bib
csl: ../References/science.csl
header-includes:
- \usepackage{amsmath}
- \usepackage{float} # Use the 'float' package
- \floatplacement{figure}{H} # Make every figure with caption = h
---

@@

The effectiveness of managing fisheries targeting shared stocks is challenged by the shifts in marine species distribution under climate change [@Pinsky:2018cb; @Pinsky:wq]. In many cases, catch or fishing effort quota on shared stocks are based on historical records, [@Baudron:2020dj] and do not necessarily consider the biogeography of the species [@FredstonHermann:2018kp], nor the effects of climate change in the fish stocks and fisheries [@PalaciosAbrantes:2020tv]. Misalignment between fisheries resources allocation and species’ distributional shifts have previously resulted in unsustainable harvest and international disputes [@Miller:2013iv; @Song:2017va; @Spijkers:2017ij], patterns that are expected to be exacerbated by worsening climate change [@Pinsky:2018cb]. With future shifts in species distributions under climate change, the challenges posed on shared stocks fisheries management will continue to increase in the $21^{st}$ century. Thus, there is a need to understand the time when climate change will affect the sharing distribution of these species as well as the intensity in order to prepare ocean governance to species on the move [@Pinsky:2018cb; @PalaciosAbrantes:2020tv; @Link:2010ei].

@@

We developed an index to evaluate range shifts in transboundary stocks under climate change (TI). This index is based on the shifts in the distribution centroid of a transboundary stock relative to the centroid of the neighboring EEZs that share this stock (Figure \@ref{fig:FigureS2}). The centroid of a transboundary stock was determined by the average latitude ($lat_{ts}$) and longitude ($lon_{ts}$) across the grid cells with the highest abundance within the neighboring EEZs sharing the stock e.g., sockeye salmon (*Oncorhynchus nerka*) that is transboundary between the United States and Canada. Therefore,

@@

where $lat_{per}$ and $lon_{per}$ are the latitudes and longitudes of the grid cells holding the $per^{th}$ percentile of the projected transboundary stock abundance. To focus on areas where transboundary stocks are more abundant and fishing activities are more likely to take place, we included grid cells where the projected stock abundance within the neighbouring EEZs sharing the stock was above the top $10^{th}$ percentile (e.g., *per* = 10%). We ran a sensitivity analysis (e.g. *per* = 20, 50, 90) to relax this parameter (Figure \@ref{fig:FigureS2}). The centroid of a EEZ was estimated using the *st* package in *R* (version 3.2.1) (Figure \@ref{fig:FigureS1}). For each ensemble member, neighboring EEZs and transboundary stock, we computed the distance between centroids assuming the earth is a perfect sphere and ignoring geographic barriers using the geosphere package in *R*;

@@

\begin{equation}
\label{eq:five}
index' = (P_c + P_v)/ToE
\end{equation}

where *P_c* and *P_v* indicate the proportion accounted for by shared species in total catch and value generated from transboundary fisheries within a country’s EEZ, respectively. Here, the proportion of total catch and value are represented as the average of 2005 to 2014 within individual EEZs according to the *Sea Around Us*. The index was then normalized to a 0 (less vulnerable) to 1 (more vulnerable) scale.

@@

In game theory, cooperation over a shared resource will more likely result in a better overall outcome for the sharing parties than operating on a non-cooperative basis [@Munro:1979dg]. However, the benefits that each player gets from a cooperative strategy must be above a minimum payoff i.e., *Threat Point* (TP) [@Sumaila:2013hv]. Here, we define TP as the minimum proportion of a shared stock within an EEZ that a country will accept to engage in cooperative management with the country’s sharing neighbor [@PalaciosAbrantes:2020vt; @Sumaila:2020tp]. Any proportion below the defined threat point of a country would result in a deflection from the joint management of the shared stock.

@@

We tested the results for normality (e.g. skewness = 1.02, kurtosis = 2.83) and performed a couple of non-parametric Kruskal–Wallis test by ranks [@Hollander:2013wz] to investigate geopolitical and ecological differences in the ToE of transboundary species. Specifically, we tested if the habitat association of transboundary species and the geographic location of EEZs would have any effect on the time of emergence of transboundary stocks. For both cases, our null hypothesis was that there were no significant differences in the time of emergence across habitat association nor EEZs.
<!-- Moreover, we implemented a multiple linear regression analysis to a series to explain the ToE of each species in each EEZ. Specifically, we looked at habitat preferences, catch trends, geopolitical and biome location and ToE of both bottom and sea surface temperature (See *Databases and species selection*). We hypothesize that all of these factors would have significant influence in the ToE of transboundary species. -->
All analyses were run using the statistical software *R version 3.5.2 (2018-12-20; Eggshell Igloo)* with the packages *data.table* [@Packagedatatable:2019uh], *janitor* [@PackagejanitorSim:2018ur], *rfishbase* [@PackagerfishbaseR:2019th], *sf* [@PackagesfSimpleF:2018vp], *sp* [@PackagespClasses:2019tn], *tidiverse* [@PackagetidyverseE:2017vq], *tidytext* [@PackagetidytextTe:2019vk], *gmt* [@PackagegmtInterfa:2017vh] and *zoo* [@PackagezooSInfr:2019tm], *ggrepel* [@PackageggrepelAut:2020to], *zeallot* [@PackagezeallotMul:2018wy], *viridis* [@PackageviridisDef:2018us], *cowplot* [@CowplotStreamlined:2019wt], *moments* [@PackagemomensMome:2015uh] and *pgirmess* [@PackagepgirmessSp:2018um]. All code is available at https://github.com/jepa/TransEmergence

@@

The median ToE of transboundary stocks varied significantly according to the geographic region of the neighboring EEZs (Kruskal-Wallis, X2 = 287.23, DF = 93, p < 0.001). Around one fifth (19%) of the EEZs have an average time of emergence across species prior to 2020 (Figure \@ref{fig:Figure2}). Notable exceptions where the ToE is relative late include Honduras (2071), New Zealand (2055), Ireland (2067) and Pitcairn (2070). The EEZs in Eastern Asia, Latin America and the Caribbean as well as Polynesia differed significantly (*p* < 0.05) from (almost) all other regions (see Table S1 for full statistics). Tropical latitudes also hold the majority of countries with high vulnerability to shifts in transboundary stocks due to climate change (Figure \@ref{fig:Figure2}). Only in very few cases like Brazil and Alaska’s and Russia’s polar EEZs, no stock will emerge between 2006 and 2100. From the top 10 more vulnerable countries, seven are located in tropical waters (Tokelau 100, the	Marshall Islands 92, Peru 88, Kiribati 80, Tuvalu 80, Papua New Guinea 78, and the Bahamas 74) while only three are outside the tropics (Estonia 86, Finland 82 and Sweden 78). We set an arbitrary threshold of  1 s.d. (67% confidence) to determine the ToE (see Methods). We tested the sensitivity of our results using a larger threshold (i.e.,  2 s.d for a 95% confidence). While the selected 1 s.d. threshold is less conservative than a 2 s.d., the distribution of results do not dramatically change between thresholds, with a difference in ToE of about 4 years (Figure S\@ref{fig:S2}).

```{r Figure2, eval = T,  echo = F, fig.width  = 13,fig.pos= "H", fig.height =11, fig.cap =  "Average time of emergence of transboundary species. A) Land polygon shows the vulnerability index of individual country’s fisheries management to shifts in transboundary species. The warmer the colour the greater the vulnerability. Exclusive Economic Zone polygon represents the average time of emergence across all species, with the cooler the colour the later the average time of emergence. The Density plot on the right shows the latitudinal number of emerging species across EEZ centroids (10 degrees latitude running mean). B) Time of emergence according to the United Nations sub regions. Whiskers represent 1.5* interquartile range. Box represents interquartile range as distance between first and third quartiles. Line represents median, and black points represent points outside of 1.5*IQR."}

@@

Our analysis suggests that most countries responsible for 75% (n = 47) of the yearly revenue generated from transboundary species between 2005 and 2010 will see their stocks emerge in average before 2040 (\@ref{fig:Figure3}). In some cases like Indonesia, Thailand and the United States, stocks will emerge ass soon as 2030. These 47 countries are also among those with the largest number of emerging stocks. Notably, nine of the top ten countries earning the most revenue from transboundary species, share a border.

```{r Figure3, eval = T,  echo = F, fig.width=4, fig.height=5,fig.pos= "H", fig.cap =  "Average time of emergence of transboundary species per fishing entity, aggregated by colour according to region. Fishing revenue on a logarithmic scale. Showing country names for the top 90th percentile. Countries marked with an asterisk have at least one neighboring border in the group."}

@@

We compared species ToE within EEZs by species habitat preference according to FishBase's classification (Figure \@ref{fig:Figure4}). We found significant differences in ToE by habitat preference (Kruskal-Wallis, $X^2$ = 286.48, *DF* = 93, *p* < 0.001). Specifically, the ToE of species with preferences for pelagic-oceanic and reef habitats was significantly different than other habitat preferences (See Table \ref{tab:TableS2}). In average, species associated with pelagic oceanic such as tunas (*Thunnus sp.*) and bathy-pelagic habitats like blue whiting (*Micromesistius poutassou*) as well as species associated to corals (e.g., Bigeye snapper	*Lutjanus	lutjanus*) will have an earliest time of emergence (Figure \@ref{fig:Figure4}). Only species associated to bathydemersal habitats such as Alaska plaice	*Pleuronectes	quadrituberculatus* and other habitats will have an average ToE later than 2025 (Figure \@ref{fig:Figure4}).

@@

<!-- We hypothesized that species environmental preferences (e.g. presence in certain biomes and habitats), climate change effects on water temperature (e.g. ToE of SST), human drivers (e.g. catch rate), and latitudinal gradient would have significant influence in the ToE of transboundary species (Table \ref{tab:TableS3}). Results from the multiple linear regression analysis show a significant relation between ToE of transboundary species and latitudinal gradient (*p* > 0.001), tropical (*p* = 0.013) and temperate (*p* = 0.002) biomes, EEZs in the Americas (*p* > 0.001), Asia (*p* > 0.001) and Oceania (*p* = 0.002), and some habitats (*p* > 0.001) (Table \ref{tab:Table1}). -->

@@

We estimated the changes in the SSR of transboundary stocks by the early (2020-2040) and the mid $21^{st}$ century (2040-2060), relative to present (1951-2005) highlighting changes below a neighbouring country's TP. Our results suggest that 87% (n = 238) of the world EEZs will see a change in SSR below the country's TP by 2030 (Figure \@ref{fig:Figure5}A). Globally, most EEZs will average a change lower than 25% by the early $21^{st}$ century which is expected to slightly increase in intensity towards the mid $21^{st}$ century (Figures \@ref{fig:FigureS4} and S5). However, this result masks larger changes in SSR for specific stocks and EEZs. For example, Guatemala’s SSR of *Panulirus gracilis* shared with Mexico is projected to go from a historical 12% to 26%, a change of over 50%. By early $21^{st}$ century, the percentage of transboundary stocks per EEZ that will see a change in SSR above the country's TP historical distribution (e.g., $\Delta{SSR} > \mu \pm \sigma$) will be relatively large (global mean of 69% of stocks in an EEZ). For 17 EEZs, 100% of their analyzed transboundary stocks will shift below the, while only six EEZs will have less than 25% stocks shifting below the TP (Figure \@ref{fig:Figure5}A-Polygons). In terms of direction of change, regions where EEZs are relatively small and have multiple borders (e.g. The Caribbean) will see a complex dynamic of changes in SSR with no clear patterns (Figure \@ref{fig:Figure1}B). Towards the mid $21^{st}$ century, changes in SSR per EEZ will not be substantial (Figure \@ref{fig:FigureS5}) relative to the early $21^{st}$ century but are expected to increase in intensity (Figure \@ref{fig:FigureS5}).

@@

In this paper we endeavor to look at the timing and intensity of changes in the sharing of transboundary species. Our findings highlight the early emergence of transboundary distribution shifts from historical variations for most of the shared stocks in the world indicating climate change is already affecting international fisheries management (Figure \@ref{fig:Figure1}). Moreover, we identify regional patterns of high climate risk to important fish stocks and associated fisheries at the present-day that will require fast responses from transboundary fisheries management (Figure \@ref{fig:Figure2} and \@ref{fig:Figure3}). Finally, we show that in many instances around the world, the incoming years will see many transboundary stocks shifting their shared proportion below acceptable levels for sharing nations potentially becoming future sources of conflict (Figure \@ref{fig:Figure5})

@@

Our results suggest that in most EEZs, more than half of transboundary species emerged in the early 2000s (Figure \@ref{fig:Figure1}). This is in alignment with previous research pointing to changes in marine catch  composition since the 1970s as evidence of climate change [@Cheung:2013gk, @Poloczanska:2016kk]. For example, in the early 2000s, Humboldt squid (*Dosidicus gigas*) substantially expanded its geographic range, reaching Washington state waters where a new fishery rapidly developed [@Pinsky:2014by; @Zeidberg:2007gs]. In the North East Atlantic, the emergence of Atlantic mackerel (*Scomber scombrus*) in Icelandic waters in 2010, led to years of unsettled international conflict between multiple fishing nations [@Spijkers:2017ij]. The large number of transboundary species emerging since early 2000s identified in this study can be partially attributed to the time of emergence of several environmental variables (e.g., temperature, oxygen) that influence the distribution of marine species [@Rodgers:2015jq; @Keller:2014jv; @Frolicher:2016dg; @Mahlstein:2011fy]. It has been estimated that since the early 2000s, SST and oxygen ($O_2$) have emerged in 79% and 23% of the global ocean, respectively, and by 2080 SST will have emerged in 90% of the global ocean while $O_2$ will have doubled under a high greenhouse gas emission scenario [@Rodgers:2015jq].

Regional differences in ToE are likely the result of environmental factors, species vulnerability, and the characteristics of political boundaries. For example, our findings underscore previous findings suggesting that tropical species are already on the move. The tropics are among the most species-rich biome [@Reygondeau:2019uh] and have potentially experienced some of the greatest levels of warming relative to natural internal variability [@IPCC:2019tn; @Roemmich:2015id]. In the tropics, marine species live close to their thermal tolerance, making them highly vulnerable to warming waters [@IPCC:2014bz]. Consequently, as a result of climate change, species have to move towards deeper or sub-tropical waters [@Poloczanska:2016kk] although recent studies suggest that in some regions species are shifting towards the equator and to shallow waters [@Clarke:2020ep]. The premise of a shift already happening is supported by observational evidence of tropical species appearing in sub-tropical biomes [@Fogarty:2017eo] and catch-data analysis showing a reduction in sub-tropical species catches in tropical waters between 1970 and 2006 [@Cheung:2013gk]. In addition to species natural traits and the environmental pressure from climate change, the tropics comprise areas with relatively small EEZs and many borders that match peaks of emerging stocks. Specifically, it seems like South East Asia, the Caribbean, and the Mediterranean harbour substantial increases in emerging stocks (Figure \@ref{fig:Figure2}).

@@

We have identified hotspot regions of climate risk for transboundary fisheries management (Figure \@ref{fig:Figure2}). The premise of our vulnerability index is that, from a management perspective, countries whose shifting transboundary stocks provide larger societal benefits and have an early ToE will be more vulnerable to shifts in SSR than their counterpart. Thus, climate risk hotspots for transboundary management happen where EEZs will see an early ToE and where transboundary stocks are important in terms of fisheries catches and revenues (Figure \@ref{fig:Figure2}). For example, Polynesia and Micronesia are the only regions where the average ToE across all stocks and EEZs already happened; 2019 or is about too 2021, respectively (Figure \@ref{fig:Figure5}). Countries in these regions are highly dependent on pelagic-oceanic (e.g.,tunas and bill fishes; [@Johnson:2017ev]) and reef-associated transboundary stocks [@Cabral:2018hc] for both food security and profits [@CisnerosMontemayor:2016gq; @Hanich:2018dh]. For example, in the Marshall Islands 16 transboundary stocks responsible for 92% and 94% of the transboundary fisheries revenue and catch, respectively, have, on average, emerged since 2018 resulting in high vulnerability index (Figure \@ref{fig:Figure5}). In this case, policy has to work against the clock and under potentially high societal pressure to address an issue that could be impacting important fisheries [@Burden:2019ij]. Currently, these fisheries operate under the Vessel Day Scheme - which takes into account shifts in stocks and catches making it a good example where a management system has been set up that is somewhat anticipatory [@Aqorau:2018bh]. However, a low vulnerability index does not necessarily translate to no-risk of international conflict or threat to stocks sustainability. While it gives countries more time to prepare for substantial changes, stocks with relatively small overall participation on catches or revenue can still be a source of conflict as portrayed by the Atlantic mackerel case [@Spijkers:2017ij, Chapter 1].

@@

<!-- $$I =(\frac{D_{A,y}}{sd(D_{A,h})} - \frac{D_{B,y}}{sd(D_{B,h})}) |  \\  +(\frac{D_{A,y}}{sd(D_{A,h})} - \frac{D_{C,y}}{sd(D_{B,h})}) | \\  +(\frac{D_{A,y}}{sd(D_{A,h})} - \frac{D_{Cy}}{sd(D_{B,h})})/N$$ -->

>>>>

---
output:
  bookdown::pdf_book:
    toc : no
    keep_tex: no
    number_sections: yes
geometry: margin=1in
editor_options:
  chunk_output_type: inline
bibliography: ../References/ToE_Ref.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{float} # Use the 'float' package
- \floatplacement{figure}{H} # Make every figure with caption = h
---

<!-- --- -->
<!-- title: Identifying Shifts in the Share Distribution of Transboundary Fish Stocks to Lead Time for Policy Response Under Climate Change -->
<!-- subtitle: Challenges to transboundary fisheries under climate change -->
<!-- author: 'Authors: Juliano Palacios-Abrantes*^1^, Thomas Frölicher^2^, Gabriel Reygondeau^13^, U. Rashid Sumaila^1^, Colette Wabnitz^14^, and William W.L. Cheung^1^' -->
<!-- date: 'Affiliations: ^1^Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada. ^2^Climate and Environmental Physics, Physics Institute, University of Bern.' -->
<!-- output: -->
<!--   # word_document:  -->
<!--     # reference_docx: ../References/template.docx -->
<!--   pdf_document: default -->
<!-- bibliography: ../References/ToE_Ref.bib -->
<!-- csl: ../References/science.csl -->
<!-- header-includes: -->
<!-- - \usepackage{amsmath} -->
<!-- - \usepackage{float} # Use the 'float' package -->
<!-- - \floatplacement{figure}{H} # Make every figure with caption = h -->
<!-- --- -->

@@

The effectiveness of managing fisheries targeting shared stocks is challenged by the shifts in marine species distribution under climate change [@Pinsky:2018cb; @Pinsky:2014by]. In many cases, catch or fishing effort quota on shared stocks are based on historical records, [@Baudron:2020dj] and do not necessarily consider the biogeography of the species [@FredstonHermann:2018kp], nor the effects of climate change in the fish stocks and fisheries [@PalaciosAbrantes:2020tv]. Misalignment between fisheries resources allocation and species’ distributional shifts have previously resulted in unsustainable harvest and international disputes [@Miller:2013iv; @Song:2017va; @Spijkers:2017ij], patterns that are expected to be exacerbated by worsening climate change [@Pinsky:2018cb]. With future shifts in species distributions under climate change, the challenges posed on shared stocks fisheries management will continue to increase in the $21^{st}$ century. Thus, there is a need to understand the time when climate change will affect the sharing distribution of these species as well as the intensity in order to prepare ocean governance to species on the move [@Pinsky:2018cb; @PalaciosAbrantes:2020tv; @Link:2010ei].

@@

We developed an index to evaluate range shifts in transboundary stocks under climate change (TI). This index is based on the shifts in the distribution centroid of a transboundary stock relative to the centroid of the neighboring EEZs that share this stock (Figure \@ref(fig:FigureS2)). The centroid of a transboundary stock was determined by the average latitude ($lat_{ts}$) and longitude ($lon_{ts}$) across the grid cells with the highest abundance within the neighboring EEZs sharing the stock e.g., sockeye salmon (*Oncorhynchus nerka*) that is transboundary between the United States and Canada. Therefore,

@@

where $lat_{per}$ and $lon_{per}$ are the latitudes and longitudes of the grid cells holding the $per^{th}$ percentile of the projected transboundary stock abundance. To focus on areas where transboundary stocks are more abundant and fishing activities are more likely to take place, we included grid cells where the projected stock abundance within the neighbouring EEZs sharing the stock was above the top $10^{th}$ percentile (e.g., *per* = 10%). We ran a sensitivity analysis (e.g. *per* = 20, 50, 90) to relax this parameter (Figure \@ref{fig:FigureS2)). The centroid of a EEZ was estimated using the *st* package in *R* (version 3.2.1) (Figure \@ref{fig:FigureS1)). For each ensemble member, neighboring EEZs and transboundary stock, we computed the distance between centroids assuming the earth is a perfect sphere and ignoring geographic barriers using the geosphere package in *R*;

@@

\begin{equation}
\label{eq:five}
index' = \frac{(P_c + P_v)}{ToE}
\end{equation}

where $P_c$ and $P_v$ indicate the proportion accounted for by shared species in total catch and value generated from transboundary fisheries within a country’s EEZ, respectively. Here, the proportion of total catch and value are represented as the average of 2005 to 2014 within individual EEZs according to the *Sea Around Us*. The index was then normalized to a 0 (less vulnerable) to 1 (more vulnerable) scale.

@@

In game theory, cooperation over a shared resource will more likely result in a better overall outcome for the sharing parties than operating on a non-cooperative basis [@Munro:1979dg]. However, the benefits that each player gets from a cooperative strategy must be above a minimum payoff i.e., "Threat Point" (TP) [@Sumaila:2013hv]. Here, we define TP as the minimum proportion of a shared stock within an EEZ that a country will accept to engage in cooperative management with the country’s sharing neighbor [@PalaciosAbrantes:2020vt; @Sumaila:2020tp]. Any proportion below the defined threat point of a country would result in a deflection from the joint management of the shared stock.

@@

We tested the results for normality (e.g. skewness = 1.02, kurtosis = 2.83) and performed a couple of non-parametric Kruskal–Wallis test by ranks [@Hollander:2013wz] to investigate geopolitical and ecological differences in the ToE of transboundary species. Specifically, we tested if the habitat association of transboundary species and the geographic location of EEZs would have any effect on the time of emergence of transboundary stocks. For both cases, our null hypothesis was that there were no significant differences in the time of emergence across habitat association nor EEZs.
<!-- Moreover, we implemented a multiple linear regression analysis to a series to explain the ToE of each species in each EEZ. Specifically, we looked at habitat preferences, catch trends, geopolitical and biome location and ToE of both bottom and sea surface temperature (See *Databases and species selection*). We hypothesize that all of these factors would have significant influence in the ToE of transboundary species. -->
All analyses were run using the statistical software R version 3.5.2 (2018-12-20; Eggshell Igloo) with the packages *data.table* [@Packagedatatable:2019uh], *janitor* [@PackagejanitorSim:2018ur], *rfishbase* [@PackagerfishbaseR:2019th], *sf* [@PackagesfSimpleF:2018vp], *sp* [@PackagespClasses:2019tn], *tidiverse* [@PackagetidyverseE:2017vq], *tidytext* [@PackagetidytextTe:2019vk], *gmt* [@PackagegmtInterfa:2017vh] and *zoo* [@PackagezooSInfr:2019tm], *ggrepel* [@PackageggrepelAut:2020to], *zeallot* [@PackagezeallotMul:2018wy], *viridis* [@PackageviridisDef:2018us], *cowplot* [@CowplotStreamlined:2019wt], *moments* [@PackagemomensMome:2015uh] and *pgirmess* [@PackagepgirmessSp:2018um]. All code is available at https://github.com/jepa/TransEmergence

@@

The median ToE of transboundary stocks varied significantly according to the geographic region of the neighboring EEZs (Kruskal-Wallis, X2 = 287.23, DF = 93, p < 0.001). Around one fifth (19%) of the EEZs have an average time of emergence across species prior to 2020 (Figure \@ref(fig:Figure2). Notable exceptions where the ToE is relative late include Honduras (2071), New Zealand (2055), Ireland (2067) and Pitcairn (2070). The EEZs in Eastern Asia, Latin America and the Caribbean as well as Polynesia differed significantly (*p* < 0.05) from (almost) all other regions (see Table S1 for full statistics). Tropical latitudes also hold the majority of countries with high vulnerability to shifts in transboundary stocks due to climate change (Figure \@ref(fig:Figure2). Only in very few cases like Brazil and Alaska’s and Russia’s polar EEZs, no stock will emerge between 2006 and 2100. From the top 10 more vulnerable countries, seven are located in tropical waters (Tokelau 100, the	Marshall Islands 92, Peru 88, Kiribati 80, Tuvalu 80, Papua New Guinea 78, and the Bahamas 74) while only three are outside the tropics (Estonia 86, Finland 82 and Sweden 78). We set an arbitrary threshold of  1 s.d. (67% confidence) to determine the ToE (see Methods). We tested the sensitivity of our results using a larger threshold (i.e.,  2 s.d for a 95% confidence). While the selected 1 s.d. threshold is less conservative than a 2 s.d., the distribution of results do not dramatically change between thresholds, with a difference in ToE of about 4 years (Figure \@ref(fig:FigureS2).

```{r Figure2, eval = T,  echo = F, fig.width  = 13,fig.pos= "H", fig.height =11, fig.cap =  "Average time of emergence of transboundary species. A) Land polygon shows the vulnerability index of individual country’s fisheries management to shifts in transboundary species. The warmer the colour the greater the vulnerability. Exclusive Economic Zone polygon represents the average time of emergence across all species, with the cooler the colour the later the average time of emergence. The Density plot on the right shows the latitudinal number of emerging species across EEZ centroids (10 degrees latitude running mean). Aqua color = no emerging stocks. B) Time of emergence according to the United Nations sub regions. Whiskers represent 1.5* interquartile range. Box represents interquartile range as distance between first and third quartiles. Line represents median, and black points represent points outside of 1.5*IQR."}

@@

Our analysis suggests that most countries responsible for 75% (n = 47) of the yearly revenue generated from transboundary species between 2005 and 2010 will see their stocks emerge in average before 2040 (Figure \@ref(fig:Figure3). In some cases like Indonesia, Thailand and the United States, stocks will emerge ass soon as 2030. These 47 countries are also among those with the largest number of emerging stocks. Notably, nine of the top ten countries earning the most revenue from transboundary species, share a border.

```{r Figure3, eval = T,  echo = F, fig.width=5, fig.height=6,fig.pos= "H", fig.cap =  "Average time of emergence of transboundary species per fishing entity, aggregated by colour according to region. Fishing revenue on a logarithmic scale. Showing country names for the top 90th percentile. Countries marked with an asterisk have at least one neighboring border in the group."}

@@

We compared species ToE within EEZs by species habitat preference according to FishBase's classification (Figure \@ref(fig:Figure4). We found significant differences in ToE by habitat preference (Kruskal-Wallis, $X^2$ = 286.48, *DF* = 93, *p* < 0.001). Specifically, the ToE of species with preferences for pelagic-oceanic and reef habitats was significantly different than other habitat preferences (See Table @\ref(tab:TableS2). In average, species associated with pelagic oceanic such as tunas (*Thunnus sp.*) and bathy-pelagic habitats like blue whiting (*Micromesistius poutassou*) as well as species associated to corals (e.g., Bigeye snapper	*Lutjanus	lutjanus*) will have an earliest time of emergence (Figure \@ref(fig:Figure4)). Only species associated to bathydemersal habitats such as Alaska plaice	*Pleuronectes	quadrituberculatus* and other habitats will have an average ToE later than 2025 (Figure \@ref{fig:Figure4)).

@@

<!-- We hypothesized that species environmental preferences (e.g. presence in certain biomes and habitats), climate change effects on water temperature (e.g. ToE of SST), human drivers (e.g. catch rate), and latitudinal gradient would have significant influence in the ToE of transboundary species (Table \ref{tab:TableS3)). Results from the multiple linear regression analysis show a significant relation between ToE of transboundary species and latitudinal gradient (*p* > 0.001), tropical (*p* = 0.013) and temperate (*p* = 0.002) biomes, EEZs in the Americas (*p* > 0.001), Asia (*p* > 0.001) and Oceania (*p* = 0.002), and some habitats (*p* > 0.001) (Table \ref{tab:Table1)). -->

@@

We estimated the changes in the SSR of transboundary stocks by the early (2020-2040) and the mid $21^{st}$ century (2040-2060), relative to present (1951-2005) highlighting changes below a neighbouring country's TP. Our results suggest that 87% (n = 238) of the world EEZs will see a change in SSR below the country's TP by 2030 (Figure \@ref{fig:Figure5}A). Globally, most EEZs will average a change lower than 25% by the early $21^{st}$ century which is expected to slightly increase in intensity towards the mid $21^{st}$ century (Figures \@ref{fig:FigureS4} and S5). However, this result masks larger changes in SSR for specific stocks and EEZs. For example, Guatemala’s SSR of *Panulirus gracilis* shared with Mexico is projected to go from a historical 12% to 26%, a change of over 50%. By early $21^{st}$ century, the percentage of transboundary stocks per EEZ that will see a change in SSR above the country's TP historical distribution (e.g., $\Delta{SSR} > \mu \pm \sigma$) will be relatively large (global mean of 69% of stocks in an EEZ). For 17 EEZs, 100% of their analyzed transboundary stocks will shift below the, while only six EEZs will have less than 25% stocks shifting below the TP (Figure \@ref{fig:Figure5}A-Polygons). In terms of direction of change, regions where EEZs are relatively small and have multiple borders (e.g. The Caribbean) will see a complex dynamic of changes in SSR with no clear patterns (Figure \@ref{fig:Figure1}B). Towards the mid $21^{st}$ century, changes in SSR per EEZ will not be substantial (Figure \@ref{fig:FigureS5)) relative to the early $21^{st}$ century but are expected to increase in intensity (Figure \@ref{fig:FigureS5)).

@@

In this paper we endeavor to look at the timing and intensity of changes in the sharing of transboundary species. Our findings highlight the early emergence of transboundary distribution shifts from historical variations for most of the shared stocks in the world indicating climate change is already affecting international fisheries management (Figure \@ref{fig:Figure1)). Moreover, we identify regional patterns of high climate risk to important fish stocks and associated fisheries at the present-day that will require fast responses from transboundary fisheries management (Figure \@ref{fig:Figure2} and \@ref{fig:Figure3)). Finally, we show that in many instances around the world, the incoming years will see many transboundary stocks shifting their shared proportion below acceptable levels for sharing nations potentially becoming future sources of conflict (Figure \@ref{fig:Figure5))

@@

Our results suggest that in most EEZs, more than half of transboundary species emerged in the early 2000s (Figure \@ref{fig:Figure1)). This is in alignment with previous research pointing to changes in marine catch  composition since the 1970s as evidence of climate change [@Cheung:2013gk, @Poloczanska:2016kk]. For example, in the early 2000s, Humboldt squid (*Dosidicus gigas*) substantially expanded its geographic range, reaching Washington state waters where a new fishery rapidly developed [@Pinsky:2014by; @Zeidberg:2007gs]. In the North East Atlantic, the emergence of Atlantic mackerel (*Scomber scombrus*) in Icelandic waters in 2010, led to years of unsettled international conflict between multiple fishing nations [@Spijkers:2017ij]. The large number of transboundary species emerging since early 2000s identified in this study can be partially attributed to the time of emergence of several environmental variables (e.g., temperature, oxygen) that influence the distribution of marine species [@Rodgers:2015jq; @Keller:2014jv; @Frolicher:2016dg; @Mahlstein:2011fy]. It has been estimated that since the early 2000s, SST and oxygen ($O_2$) have emerged in 79% and 23% of the global ocean, respectively, and by 2080 SST will have emerged in 90% of the global ocean while $O_2$ will have doubled under a high greenhouse gas emission scenario [@Rodgers:2015jq].

Regional differences in ToE are likely the result of environmental factors, species vulnerability, and the characteristics of political boundaries. For example, our findings underscore previous findings suggesting that tropical species are already on the move. The tropics are among the most species-rich biome [@Reygondeau:2019uh] and have potentially experienced some of the greatest levels of warming relative to natural internal variability [@IPCC:2019tn; @Roemmich:2015id]. In the tropics, marine species live close to their thermal tolerance, making them highly vulnerable to warming waters [@IPCC:2014bz]. Consequently, as a result of climate change, species have to move towards deeper or sub-tropical waters [@Poloczanska:2016kk] although recent studies suggest that in some regions species are shifting towards the equator and to shallow waters [@Clarke:2020ep]. The premise of a shift already happening is supported by observational evidence of tropical species appearing in sub-tropical biomes [@Fogarty:2017eo] and catch-data analysis showing a reduction in sub-tropical species catches in tropical waters between 1970 and 2006 [@Cheung:2013gk]. In addition to species natural traits and the environmental pressure from climate change, the tropics comprise areas with relatively small EEZs and many borders that match peaks of emerging stocks. Specifically, it seems like South East Asia, the Caribbean, and the Mediterranean harbour substantial increases in emerging stocks (Figure \@ref{fig:Figure2)).

@@

We have identified hotspot regions of climate risk for transboundary fisheries management (Figure \@ref{fig:Figure2)). The premise of our vulnerability index is that, from a management perspective, countries whose shifting transboundary stocks provide larger societal benefits and have an early ToE will be more vulnerable to shifts in SSR than their counterpart. Thus, climate risk hotspots for transboundary management happen where EEZs will see an early ToE and where transboundary stocks are important in terms of fisheries catches and revenues (Figure \@ref{fig:Figure2)). For example, Polynesia and Micronesia are the only regions where the average ToE across all stocks and EEZs already happened; 2019 or is about too 2021, respectively (Figure \@ref{fig:Figure5)). Countries in these regions are highly dependent on pelagic-oceanic (e.g.,tunas and bill fishes; [@Johnson:2017ev]) and reef-associated transboundary stocks [@Cabral:2018hc] for both food security and profits [@CisnerosMontemayor:2016gq; @Hanich:2018dh]. For example, in the Marshall Islands 16 transboundary stocks responsible for 92% and 94% of the transboundary fisheries revenue and catch, respectively, have, on average, emerged since 2018 resulting in high vulnerability index (Figure \@ref{fig:Figure5)). In this case, policy has to work against the clock and under potentially high societal pressure to address an issue that could be impacting important fisheries [@Burden:2019ij]. Currently, these fisheries operate under the Vessel Day Scheme - which takes into account shifts in stocks and catches making it a good example where a management system has been set up that is somewhat anticipatory [@Aqorau:2018bh]. However, a low vulnerability index does not necessarily translate to no-risk of international conflict or threat to stocks sustainability. While it gives countries more time to prepare for substantial changes, stocks with relatively small overall participation on catches or revenue can still be a source of conflict as portrayed by the Atlantic mackerel case [@Spijkers:2017ij, Chapter 1].

@@

<!-- $$I =(\frac{D_{A,y}}{sd(D_{A,h))} - \frac{D_{B,y}}{sd(D_{B,h)))) |  \\  +(\frac{D_{A,y}}{sd(D_{A,h))} - \frac{D_{C,y}}{sd(D_{B,h)))) | \\  +(\frac{D_{A,y}}{sd(D_{A,h))} - \frac{D_{Cy}}{sd(D_{B,h))))/N$$ -->
